The role of standing variation in parallel adaptation

### 2014-09-02 standing-variation-paper.tex
added char length as function of sb to R script, and edits to man text

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\newcommand{\gc}[1]{{\it\color{green}(#1)} }
\newcommand{\plr}[1]{{\it\color{blue}(#1)}}

@@

%\iflatexml  % no margin notes in latexml
% \newcommand{\marginnote}[1]{}
%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
%\fi

\bibliographystyle{plain}

@@

@@

We assume that dispersal occurs by local movement of
individuals. Specifically we will assume that individuals dispersal a
normally distributed distance, with the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance)

@@

\paragraph{Geographic spread of alleles} 
Once an allele has become local established it can begin to spatially spread. 
We assume that the allele, having become established, quickly settles
down to spread spatially as a traveling wave of constant speed. The behavior of
this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}. Under reasonably general
conditions the speed of advance
of this wave will be $v = \sigma \sqrt{2s}$. See
\citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
applies to this scenerio. Note that the speed of the wave can be
changed

@@

Our results are predicated on the idea that adaptive variants are
initially rare within populations, i.e. they are reasonably
deleterious before the environment switches. In constrast, even distant populations could have
a shared basis of adaptation if they adapt via common variation,
e.g. previously neutral (or nearly neutral) variation, shared among populations. 
Our ability to detect such shared convergent events is at the moment
quite limited. 
One way that selection pressures might be weak on alleles, is if they
are at only one of many loci contributing to variation in a trait. However, if that
case populations separated by a reasonable amount of drift may still
adapt via different genetic routes, as the constellation of alleles
best positioned to initially respond to positive selection, i.e. those at
intermediate frequency, will be somewhat different.

@@

\paragraph{How then do substitutions occur?}
It is rare for gene flow to rapidly spread selected alleles across
species range, how then do selected alleles ever become fixed within
species? Repeated bouts of adaptation in particular genomic regions may act 
push a subset of previously selected alleles to fixation across the
species range, through the spread of the genetic backgrounds on which they arise.  
However, it seems likely that this is a slow process compared to the
initial rapid spread of selected alleles.

@@

Somewhat along similar lines, Futuyma has argued that much of the adaptive differentiation,
e.g. adaptation to local conditions, within species maybe ephemeral and
subject to loss due to local extinction and the mixing following the collapse of population
structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
response to punctuated equilibrium \citep{eldredgegould72}, arguing
that the observation of stasis and rapid anagensis associated with
speciation were consistent with micro-evolution . 
Futuyma argued that despite rapid adaptation over short time-scales, we
may observe morphological stasis in the
fossil record as
much of this adaptation is lost \citep[see
also][]{Eldredge:05}. Furthermore, he suggests that speciation
may act as rachet to prevent the loss of differentiation, acting to
maintain adaptive changes among populations, and prevent their loss by
interbreeding. At face value the rate of species formation seems too low to
contribute to this process. However, as \citep{Rosenblum:12}, and many
others, have argued that in fact the rate of speciation may well be very high,
but that the majority of incipient species do not persist long due to
reabsorption or extinction. Repeated bouts of extinction and
speciation will send waves of alleles
to fixation along particular lineages.

This link between speciation and substitution does not imply that
substitutions should be thought of as being clustered at splits in
inferred phylogenies \citep[see ][for a recent exhange on
this]{Venditti:10, Pennell:14,Venditti:14,Pennell:14B}. 
(Note that neutral substitutions will be to be unaffected by this process, as they
will accumulate in a clock like manner along lineages, as dictated by
the mutation rate, regardless of the geographic details of their
polymorphic stage.)
Firstly, it it likely that splits in
phylogenies are likely only a tiny proportion of all incipient
speciation events because extinction rates may be high
\citep{Rosenblum:12}, such that every lineage has likely past through
many ``speciation events'' in addition to the observed ones.
Secondly, under this view, the origination process of alleles and their fixation process
are uncoupled \citep[in the sense of ][ page XXX]{Gillespiebook} by the geographic structure.
Such that spatial polymorphisms may accumulate gradually, in
geographically restricted populations, between large-scale 
biogeographic events. 
%Thus selected alleles that eventually become a
%substitutions may have arisen long before they reach fixation.

Nor does this view imply that adaptive evolution or speciation is driven by the
shifting balance or genetic revolutions \citep{Wright:32,
  Mayr-genetic-revol:1954}, whereby extreme genetic drift allows
populations to cross fitness valleys and substitute novel epistatic
combinations. While geographic lineage sorting via speciation and
extinction can be thought of as very large-scale genetic drift events,
the initial spread of our alleles is due to selection they are only later sorted

This view suggests that the alleles that substitute along a lineage
may not be those but rather are in part those lucky enough to have
spread across areas not

contributing to

@@

species range evolution
splitting of species.

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[authoryear]{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\newcommand{\gc}[1]{{\it\color{green}(#1)} }
\newcommand{\plr}[1]{{\it\color{blue}(#1)}}

@@

%\iflatexml  % no margin notes in latexml
% \newcommand{\marginnote}[1]{}
%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
%\fi
\bibliographystyle{amnat}
%\bibliographystyle{plain}

\title{The role of standing variation in parallel adaptation}
\author{Peter L. Ralph$^1$ \\ email: pralph@usc.edu  \and Graham Coop$^2$ \\ email: gmcoop@ucdavis.edu }

@@

\maketitle
\date{}

\begin{center}

$^1$ Computational Biology and Bioinformatics \\ 
University of Southern California, Los Angeles, CA \\
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ 
University of California -- Davis, Davis, CA, 95616
\end{center}

\section*{Abstract}

@@

%We assume that dispersal occurs by local movement of
%individuals. Specifically we will assume that individuals dispersal a
%normally distributed distance, with the mean squared distance between parent and child is $\sigma^2$
%(i.e.\ $\sigma$ is the dispersal distance)

@@

\paragraph{Geographic spread of alleles} 
Once an allele has become local established it can begin to spatially spread. 
We assume that the allele, having become established, quickly settles
down to spread spatially as a traveling wave of constant speed. The behavior of
this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}. Under reasonably general
conditions the speed of advance
of this wave will be $v = \sigma \sqrt{2s}$. See
\citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
applies to this scenerio. Note that the speed of the wave will vary
with details of the space that individuals migrate across
\citep[e.g. see ][ for comparisons to migration on discrete
grids]{Slatkin-speed:76,SlatkinCharlesworth:78}, however, the velocity
will usually be primarily determined by the product $\sigma^2$ and $s$.

@@

\paragraph{}
Our results are predicated on the idea that adaptive variants are
initially rare within populations, i.e. they are reasonably
deleterious before the environment switches. In constrast, even distant populations could have
a shared basis of adaptation if they adapt via common variation,
e.g. previously neutral (or nearly neutral) variation, shared among populations. 
Our ability to detect such shared convergent events is at the moment
quite limited. If many loci contributing to variation in a trait, then
selection on any one allele may be weak allowing them to segregate
across populations. However, if that
case populations separated by a reasonable amount of drift may still
adapt via different genetic routes, as the constellation of alleles
best positioned to initially respond to positive selection, i.e. those at
intermediate frequency, will be somewhat different. Therefore, it
seems unlikely that polygenic traits are more suspectible to a shared
genetic basis to adaptation across populations than simple traits.

@@

\paragraph{How then do substitutions occur?}
If it is rare for gene flow to rapidly spread selected alleles across
species range, how then do selected alleles ever become fixed within
species? Repeated bouts of adaptation in particular genomic regions may act 
push a subset of previously selected alleles to fixation across the
species range, through the spread of the genetic backgrounds on which they arise.  
However, it seems likely that this is a slow process compared to the
initial rapid spread of selected alleles.

@@

Somewhat along similar lines, Futuyma has argued that much of the adaptive differentiation,
e.g. adaptation to local conditions, within species maybe ephemeral and
subject to loss due to local extinction and the mixing following the collapse of population
structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
response to punctuated equilibrium \citep{eldredgegould72}, arguing
that the observation of stasis and rapid anagensis associated with
speciation were consistent with micro-evolution. 
Futuyma argued that despite rapid adaptation over short time-scales, we
may observe morphological stasis in the
fossil record as
much of this adaptation is lost \citep[see
also][]{Eldredge:05}. Furthermore, he suggests that speciation
may act as rachet to prevent the loss of differentiation, acting to
maintain adaptive changes among populations, and prevent their loss by
interbreeding. At face value the rate of species formation seems too low to
contribute to this process. However, as \citep{Rosenblum:12}, and many
others, have argued that in fact the rate of speciation may well be very high,
but that the majority of incipient species do not persist long due to
reabsorption or extinction. Repeated bouts of extinction and
speciation will send waves of alleles
to fixation along particular lineages.

%Venditti:10, 
This link between speciation and substitution does not imply that
substitutions should be thought of as being clustered at splits in
inferred phylogenies \citep[see ][for a recent exhange on
this]{Pennell:14,Venditti:14,Pennell:14B}. 
(Note that neutral substitutions will be to be unaffected by this process, as they
will accumulate in a clock like manner along lineages, as dictated by
the mutation rate, regardless of the geographic details of their
polymorphic stage.)
First, it it likely that splits in
phylogenies are likely only a tiny proportion of all incipient
speciation events because extinction rates may be high
\citep{Rosenblum:12}, such that every lineage has likely past through
many ``speciation events'' in addition to the observed ones.
Second, under this view, spatial polymorphisms may accumulate gradually, in
geographically restricted populations, between large-scale 
biogeographic events leading to their fixation (or indeed loss). 
Such that the origination process of alleles and their fixation process
are reasonably uncoupled \citep[in the sense of ][]{Gillespie:94} by the geographic structure.

%Thus selected alleles that eventually become a
%substitutions may have arisen long before they reach fixation.

Nor does this view imply that adaptive evolution or speciation is driven by the
shifting balance or genetic revolutions \citep{Wright:32,
  Mayr-genetic-revol:1954}, whereby genetic drift allows
populations to cross fitness valleys and substitute novel epistatic
combinations. While geographic lineage sorting via speciation and
extinction can be thought of as very large-scale genetic drift events,
the initial spread of our alleles is due to selection they are only later sorted 
into substitutions via these large scale drift events \citep[see also
][for discussion]{Futuyma:89}.

This view suggests that the alleles that substitute along a lineage
may rarely be capable, or need, to spread species-wide. This is not
inconsistent with genome-wide evidence that a reasonable fraction of 
substitutions are fixed by positive selection in a number of species
\citep[most notably Drosophila][]{}. Selection has played a strong
role in the establishment of these alleles, and their frequency may be
high enough within populations that they do not contribute
substantially to polymorphism.

@@
